Automatic conversion of colloquial Finnish to standard Finnish

### 2015-01-26 finnormal-nodalida-2015/finnormal2015.tex
cleanup
git-svn-id: https://svn.code.sf.net/p/apertium/svn/branches/papers/@58579 72bbbca6-d526-0410-a7d9-f06f51895060

\cite{viinikka2013} describe the common meaning of the terms colloquial (\emph{puhekieli}) and standard (\emph{yleiskieli} or \emph{kirjakieli}) Finnish: standard language follows the recommendations of kielenhuolto and is unified in morphology and vocabulary; colloquial language shows local and idiolectal variation, and has structures that are characteristic to spoken variety,  such as discourse particles and incomplete clauses.
Prototypically, colloquial language is spoken, everyday and spontaneous, and standard is written, planned and structured.

@@

Morphology: dialectal variation, shortening and merging of morphemes, e.g. \emph{en min\"{a}} `not I' $\rightarrow$ \emph{emm\"{a}} (`I don't').

Syntax: due to the spontaneous nature of colloquial language, syntactic structures are common for spoken language. The following example is from our data set, first sentence is the original colloquial version. The gloss shows the actual word-by-word translation, and the translation shows a similar style and register in English.

@@

Sentence 2 is translated in many ways. On the lexical level, the word \emph{seiska} is a colloquial name for `number 7', and in a literary standard it is replaced by the ordinal \emph{seitsemäs} `seventh'. The syntactic structure of the sentence is markedly spoken; the word \emph{seiska} is topicalised, and the main information ``deduced from phonotactics'' is a subordinate clause. 
The translation below is shorter and more precise, leaving just the main information.

@@

For the partial version, we didn't touch the syntactic structure, but changed morphology and vocabulary. Below is a version that shows the original and the partial.

\ex
\begingl
\gla seiskakin oli vaan silleen et fonotaksista p\"{a}\"{a}ttelin //
\glb seitsemäskin oli vain sill\"{a}\#lailla ett\"{a} fonotaksista p\"{a}\"{a}ttelin //
%\glb seventh-{\sc also} was just like.that that phonotactics.{\sc ela} I.deduced //
% \glft `also the seventh, it was like, I just deduced it from phonotactics' //
\endgl
\xe

>>>>

\cite{viinikka2013} describe the common meaning of the terms colloquial (\emph{puhekieli}) and standard (\emph{yleiskieli} or \emph{kirjakieli}) Finnish: standard language is regulated by kielenhuolto and is unified in morphology and vocabulary; colloquial language shows local and idiolectal variation, and has structures that are characteristic to spoken variety,  such as discourse particles and incomplete clauses.

@@

Morphology shows dialectal variation, shortening and merging of morphemes, e.g. \emph{en min\"{a}} `not I' $\rightarrow$ \emph{emm\"{a}} (`I don't').

Syntax: due to the spontaneous nature of colloquial language, syntactic structures are common for spoken language. 
Prototypically, colloquial language is spoken and spontaneous, and standard is written, planned and structured. We illustrate the differences with the following example from our data set. Sentence 1 is the original colloquial version. The gloss shows the actual word-by-word translation, and the translation shows similar style and register in English.

@@

For the partial version, we didn't touch the syntactic structure, but changed morphology and vocabulary. On the lexical level, the word \emph{seiska} `number 7' is colloquial style, and in the standard translation it is replaced by the ordinal \emph{seitsem\"{a}s} `seventh'.  Other changes in the partial version are common morphological or phonological, phenomena, such as reducing the diphthong in \emph{vaan} $\rightarrow$ \emph{vain} or shortening in  \emph{et} $\rightarrow$ \emph{että}. The original sentence and the partial translation are shown below, aligned word by word.

\ex
\begingl
\gla seiskakin oli vaan silleen et fonotaksista p\"{a}\"{a}ttelin //
\glb seitsem\"{a}skin oli vain sill\"{a}\#lailla ett\"{a} fonotaksista p\"{a}\"{a}ttelin //
%\glb seventh-{\sc also} was just like.that that phonotactics.{\sc ela} I.deduced //
% \glft `also the seventh, it was like, I just deduced it from phonotactics' //
\endgl
\xe

 The syntactic structure of the original sentence is markedly spoken; the word \emph{seiska} is topicalised, and the main information ``deduced from phonotactics'' is in a subordinate clause. 
The translation into  is shorter and more precise, leaving just the main information.

@@
